---
title: Multi-sample analyses
vignette: >
  % \VignetteIndexEntry{Multi-sample analysis}
  %\VignetteEncoding{UTF-8}
  % \VignetteEngine{knitr::rmarkdown}
output:
  html_document:
    mathjax: null   
editor_options: 
  chunk_output_type: console
---


# Setup and data exploration

As said, we will use the the wild-type data from the Tal1 chimera experiment:

- Sample 5: E8.5 injected cells (tomato positive), pool 3
- Sample 6: E8.5 host cells (tomato negative), pool 3
- Sample 7: E8.5 injected cells (tomato positive), pool 4
- Sample 8: E8.5 host cells (tomato negative), pool 4
- Sample 9: E8.5 injected cells (tomato positive), pool 5
- Sample 10: E8.5 host cells (tomato negative), pool 5

Note that this is a paired design in which for each biological replicate (pool 3, 4, and 5), we have both host and injected cells.

We start by loading the data and doing a quick exploratory analysis, essentially applying the normalization and visualization techniques that we have seen in the previous lectures to all samples.

```{r chunk-opts, include=FALSE}
rm(list = ls())
gc()
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message=FALSE, warning=FALSE)
library(BiocStyle)
```


```{r setup, message = FALSE}
library(MouseGastrulationData)
sce <- WTChimeraData(type = "processed", samples=5:10)
sce
colData(sce)
```

To speed up computations, after removing doublets, we randomly select 50% cells per sample.

```{r}
library(scater)
library(ggplot2)
library(scran)

# remove doublets
drop <- sce$celltype.mapped %in% c("stripped", "Doublet")
sce <- sce[,!drop]

set.seed(29482)
idx <- unlist(tapply(colnames(sce), sce$sample, function(x) {
    perc <- round(0.50 * length(x))
    sample(x, perc)
}))

sce <- sce[,idx]
```

We now normalize the data and visualize them in a tSNE plot.

```{r}
# normalization
sce <- logNormCounts(sce)

# identify highly variable genes
dec <- modelGeneVar(sce, block=sce$sample)
chosen.hvgs <- dec$bio > 0

# dimensionality reduction
sce <- runPCA(sce, subset_row = chosen.hvgs, ntop = 1000)
sce <- runTSNE(sce, dimred = "PCA")

sce$sample <- as.factor(sce$sample)
plotTSNE(sce, colour_by = "sample")
plotTSNE(sce, colour_by = "celltype.mapped") +
    scale_color_discrete() +
    theme(legend.position = "bottom")
```

There are evident sample effects. Depending on the analysis that you want to perform you may want to remove or retain the sample effect. For instance, if the goal is to identify cell types with a clustering method, one may want to remove the sample effects with "batch effect" correction methods.

For now, let's assume that we want to remove this effect.

# Correcting batch effects

We correct the effect of samples by aid of the `correctExperiment` function
in the `batchelor` package and using the `sample` `colData` column as batch.


```{r}

library(batchelor)
set.seed(10102)
merged <- correctExperiments(sce, 
    batch=sce$sample, 
    subset.row=chosen.hvgs,
    PARAM=FastMnnParam(
        merge.order=list(
            list(1,3,5), # WT (3 replicates)
            list(2,4,6)  # td-Tomato (3 replicates)
        )
    )
)

merged <- runTSNE(merged, dimred="corrected")
plotTSNE(merged, colour_by="batch")

```

Once we removed the sample batch effect, we can proceed with the Differential 
Expression Analysis.


# Differential Expression

In order to perform a Differential Expression Analysis, we need to identify 
groups of cells across samples/conditions (depending on the experimental 
design and the final aim of the experiment). 

As previously seen, we have two ways of grouping cells, cell clustering and cell
labeling.
In our case we will focus on this second aspect to group cells according to the 
already annotated cell types to proceed with the computation of the 
pseudo-bulk samples.

## Pseudo-bulk samples

To compute differences between groups of cells, a possible way is to 
compute pseudo-bulk samples, where we mediate the gene signal of all the cells
for each specific cell type.
In this manner, we are then able to detect differences between the same cell type 
across two different conditions.

To compute pseudo-bulk samples, we use the `aggregateAcrossCells` function in the 
`scuttle` package, which takes as input not only a SingleCellExperiment, 
but also the id to use for the identification of the group of cells.
In our case, we use as id not just the cell type, but also the sample, because
we want be able to discern between replicates and conditions during further steps.

```{r}
# Using 'label' and 'sample' as our two factors; each column of the output
# corresponds to one unique combination of these two factors.
library(scuttle)
summed <- aggregateAcrossCells(merged, 
    id=colData(merged)[,c("celltype.mapped", "sample")])
summed

```

## Differential Expression Analysis

The main advantage of using pseudo-bulk samples is the possibility to use 
well-tested methods for differential analysis like `edgeR` and `DESeq2`, we will
focus on the former for this analysis.
`edgeR` uses a Negative Binomial Generalized Linear Model.

First, let's start with a specific cell type, for instance the "Mesenchymal stem cells", and look into differences between this cell type across conditions.

```{r}
label <- "Mesenchyme"
current <- summed[,label==summed$celltype.mapped]

# Creating up a DGEList object for use in edgeR:
library(edgeR)
y <- DGEList(counts(current), samples=colData(current))
y
```

A typical step is to discard low quality samples due to low sequenced library 
size. We discard these samples because they can affect further steps like normalization
and/or DEGs analysis.

We can see that in our case we don't have low quality samples and we don't need 
to filter out any of them.

```{r}
discarded <- current$ncells < 10
y <- y[,!discarded]
summary(discarded)
```

The same idea is typically applied to the genes, indeed we need to discard low 
expressed genes to improve accuracy for the DEGs modeling.

```{r}
keep <- filterByExpr(y, group=current$tomato)
y <- y[keep,]
summary(keep)
```

We can now proceed to normalize the data
There are several approaches for normalizing bulk, and hence pseudo-bulk data. Here, we use the Trimmed Mean of M-values method, implemented in the `edgeR` 
package within the `calcNormFactors` function.
Keep in mind that because we are going to normalize the pseudo-bulk counts,
we don't need to normalize the data in "single cell form".

```{r}
y <- calcNormFactors(y)
y$samples
```

To investigate the effect of our normalization, we use a Mean-Difference (MD) plot for each sample
in order to detect possible normalization problems due to insufficient cells/reads/UMIs composing a particular pseudo-bulk profile.

In our case, we verify that all these plots are centered in 0 (on y-axis) and present
a trumpet shape, as expected.


```{r}
par(mfrow=c(2,3))
for (i in seq_len(ncol(y))) {
    plotMD(y, column=i)
}
```

Furthermore, we want to check if the samples cluster together based
on their known factors (like the tomato injection in this case).

To do so, we use the MDS plot, which is very close to a PCA representation.
    
```{r}
par(mfrow=c(1,1))
limma::plotMDS(cpm(y, log=TRUE), 
    col=ifelse(y$samples$tomato, "red", "blue"))
```

We then construct a design matrix by including both the pool and the tomato as factors.
This design indicates which samples belong to which pool and condition, so we can
use it in the next step of the analysis.

```{r}
design <- model.matrix(~factor(pool) + factor(tomato), y$samples)
design
```

Now we can estimate the Negative Binomial (NB) overdispersion parameter, to model
the mean-variance trend.

```{r}
y <- estimateDisp(y, design)
summary(y$trended.dispersion)
```

We then fit a Quasi-Likelihood (QL) negative binomial generalized linear model for each gene. 
The `robust=TRUE` parameter avoids distortions from highly variable clusters.
The QL method includes an additional dispersion parameter, useful to handle the uncertainty and variability of the per-gene variance, which is not well estimated by the NB dispersion, so the two dispersion types complement each other in the final analysis.

```{r}
fit <- glmQLFit(y, design, robust=TRUE)
summary(fit$var.prior)
summary(fit$df.prior)
```

We then use an empirical Bayes quasi-likelihood F-test to test for differential expression (due to tomato injection) per each gene at a False Discovery Rate (FDR) of 5%.
The low amount of DGEs highlights that the tomato injection effect has a low 
influence on the mesenchyme cells.

```{r}
res <- glmQLFTest(fit, coef=ncol(design))
summary(decideTests(res))
topTags(res)
```

All the previous steps can be easily performed with the following function 
for each cell type, thanks to the `pseudoBulkDGE` function in the `scran` package.

```{r}
library(scran)
summed.filt <- summed[,summed$ncells >= 10]

de.results <- pseudoBulkDGE(summed.filt, 
    label=summed.filt$celltype.mapped,
    design=~factor(pool) + tomato,
    coef="tomatoTRUE",
    condition=summed.filt$tomato 
)
```

The returned object is a list of `DataFrame`s each with the results for a cell type.
Each of these contains also the intermediate results in `edgeR` format to perform any intermediate plot or diagnostic.

```{r}
cur.results <- de.results[["Allantois"]]
cur.results[order(cur.results$PValue),]

```


# Differential Abundance

With DA we test for differences between clusters across conditions, to investigate
which clusters change accordingly to the treatment (the tomato injection in our case).

We first setup some code and variables for further analysis, like quantifying the
number of cells per each cell type and fit a model to catch differences between the 
injected cells and the background.

The performed steps are very similar to the ones for DEGs analysis, but this time
we start our analysis on the computed abundances and without normalizing the 
data with TMM.


```{r}
library(RColorBrewer)
n <- 68
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

abundances <- table(merged$celltype.mapped, merged$sample) 
abundances <- unclass(abundances) 
abundances

df <- as.data.frame(abundances)
df <- cbind(rowSums(df[,c(1, 3, 5)]), rowSums(df[, c(2, 4, 6)]))
colnames(df) <- c("tomato", "control")

# Create a df for barplot
ct <- rep(rownames(df), 2)
cond <- c(rep("tomato", dim(df)[1]), rep("control",dim(df)[1]))
value <- c(df[,1], df[,2])

df_barplot <- data.frame(ct, cond, value)


ggplot(df_barplot, aes(x=reorder(ct, -value), y=value, fill=cond))+ 
  geom_bar(width=.5, stat='identity', position=position_dodge(.7)) + 
    theme_classic() +theme(axis.text.x = element_text(angle = 75, vjust = 1, hjust=1))+
    labs(y="#Cells", x="")

# Barplot
ggplot(df_barplot, aes(y=value, x=cond, fill=ct, color=ct)) + 
  geom_bar(position="fill", stat="identity") + 
  scale_color_manual(values=col_vector)+
  scale_fill_manual(values=col_vector) +
    labs(y="Cell proportion(%)", x="", fill=NULL, color=NULL) +theme_classic() 

```

```{r}
library(edgeR)
abundances <- table(merged$celltype.mapped, merged$sample) 
abundances <- unclass(abundances) 
# Attaching some column metadata.
extra.info <- colData(merged)[match(colnames(abundances), merged$sample),]
y.ab <- DGEList(abundances, samples=extra.info)
keep <- filterByExpr(y.ab, group=extra.info$tomato)
y.ab <- y.ab[keep,]
design <- model.matrix(~factor(pool) + factor(tomato), y.ab$samples)
y.ab <- estimateDisp(y.ab, design, trend="none")
fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
```

## Background on compositional effect

As mentioned before, in DA we don't normalize our data with `calcNormFactors` 
function, because this approach considers that most of the input features do not vary between conditions.
This cannot be applied to this analysis because we have a small number of cell
populations that all can change due to the treatment.
Leaving us to normalize only for library depth, which in pseudo-bulk data means 
by the total number of cells in each sample (cell type).

On the other hand, this can lead our data to be susceptible to compositional 
effect, that means that our conclusions can be biased by the amount of cells 
present in each cell type. And this amount of cells can be totally unbalanced 
between cell types.

For example, a specific cell type can be 40% of the total amount of cells
present in the experiment, while another just the 3%. 
The differences in terms of abundance of these cell types are detected between 
the different conditions, but our final interpretation could be biased if we don't
consider this aspect.

We now look at one of the approaches for handling the compositional effect.

## Assuming most labels do not change

We can use a similar approach used during the DEGs analysis, assuming that most
labels are not changing, in particular if we think about the low number of DEGs 
resulted from the previous analysis.

To do so, we first normalize the data with `calcNormFactors` and then we fit and 
estimate a QL-model for our abundance data.

```{r}
y.ab2 <- calcNormFactors(y.ab)
y.ab2$samples$norm.factors
```

We then follow the already seen edgeR analysis steps. 

```{r}
y.ab2 <- estimateDisp(y.ab2, design, trend="none")
fit.ab2 <- glmQLFit(y.ab2, design, robust=TRUE, abundance.trend=FALSE)
res2 <- glmQLFTest(fit.ab2, coef=ncol(design))
summary(decideTests(res2))
topTags(res2, n=10)
```

# Session Info

```{r sessioninfo, tidy=TRUE}
sessionInfo()
```


# Further Reading

* OSCA book, Multi-sample analysis, [Chapters 1, 4, and 6](https://bioconductor.org/books/release/OSCA.multisample)

# Exercises

* Test differential expressed genes with just 2 replicates per condition and look into the changes in the results and possible emerging issues.

* Test differential expressed genes computed with `muscat` package and check for differences in the results.


